
---
title: "An analysis of socioeconomic factors and fire inspection violations in Toronto highrises"
author: "Jessica Im"
date: "today"
date-format: long
format: pdf
toc: TRUE
abstract: "Exploring fire investigation results and social and economic factors in Toronto."
bibliography: fire_violation_references.bib
---

```{r}
#| message: false
#| echo: false

#### Workspace setup ####

## Load packages
library(opendatatoronto)
library(dplyr)
library(tidyverse)
library(knitr)
library(janitor)
library(lubridate)
library(ggplot2)
library(readxl)

# Read in cleaned fire data
cleaned_fire_inspection_data <- 
  read.csv(here::here("inputs/data/highrise_fire_inspection_clean_all.csv"))
show_col_types = FALSE

cleaned_fire_inspection_data_ward <- 
  read.csv(here::here("inputs/data/highrise_fire_inspection_clean_ward.csv"))
show_col_types = FALSE

# Read in cleaned ward data
ward_data_cleaned <- 
  read.csv(here::here("inputs/data/ward_data_cleaned.csv"))
show_col_types = FALSE

ward_data_cleaned_socials <- 
  read.csv(here::here("inputs/data/ward_data_socials_cleaned.csv"))
show_col_types = FALSE

# Sample avg rent ward data
cleaned_ward_data_sample <-
  ward_data_cleaned[-c(7:27)]

# Sample elderly residents ward data
cleaned_socials_ward_data_sample <-
  ward_data_cleaned_socials[-c(7:27)]
```

## Introduction
The Ontario Fire Code is a regulation under the Fire Protection and Prevention Act of 1997, and helps guide Toronto Fire Services and building fire safety requirements [@fireinspectionresultswebsite]. Since 2016, TFS Fire Inspectors conduct fire safety inspections on residential highrises yearly [@firepreventioninhighrises]. Results of these fire inspections can be found on OpenDataToronto, making fire safety accessible to the public. 

According to Statistics Canada, residential fires accounted for 3 of 4 fire-related deaths. Homes without a functional smoke alarm also accounted for 74% of deaths [@fireincidentspandemic]. Unfortunately, from 2015 to 2021, only 37% of residential fires had a working smoke alarm. Fire-related injuries were also higher in homes with a functioning smoke alarm but no sprinkler protection. Statistics Canada explains this may be due to attempts by the residents to control the fires. Since the Covid-19 pandemic, fires have increased and the rise in deaths among seniors was highest; from 2019 to 2020 fire-related deaths among seniors doubled. In a cross-sectional population-based study in British Columbia, Alberta, Manitoba, and Ontario,  socioeconomic factors related to residential fire rates were investigated [@socioeconomicfactorsandfireincidents]. The report found an association between higher educational attainment and unemployment with increased fire incidents. Smaller households were also found to be at higher risk for fire casualties, especially for children. 

While high-rises are designed with fire safety as a priority, because of the number of residents and size, a fire has potential for major incidents and can be challenging to respond to [@highrisefiresafety]. This paper will examine the relationship between fire regulation violations in highrises, which requirements failed to be met, and various socio-economic factors by walking through the following sections: Data, Results, Discussion, and Conclusion. In Data, I will discuss the spreadsheets on fire inspection violations in highrises and socio-economic information by Ward Profiles. I also detail the cleaning and analyzing process of these datasets. Results will reveal trends found in the analyses and the Discussion section highlights additional insights. Finally, in the Conclusion, I will summarize key takeaways.

## Data
The two datasets used in this paper are from the city of Toronto's OpenDataToronto database [@R-opendatatoronto]. To investigate the various socioeconomic variables and investigation violations in relation to highrises, two datasets were used: "Highrise Residential Fire Inspection Results" [@FireInspectionResultsData] and "Ward Profiles (25-ward model) [@WardProfiles]. The data were cleaned and analyzed using open-source statistical programming language R [@R-base], alongside dplyr [@R-dplyr], tidyverse [@R-tidyverse], forcats [@R-forcats], ggplot2 [@R-ggplot2], janitor [@R-janitor], knitr [@R-knitr], lubridate [@R-lubridate], purrr [@R-purrr], readr [@R-readr], readxl [@R-readxl], stringr [@R-stringr], tibble [@R-tibble], and tidyr [@R-tidyr].

### Fire Inspection Results Data
The Highrise Residential Fire Inspection Results Data [@FireInspectionResultsData] consists of properties with and without identified Ontario Fire Code violations. This dataset does not include ongoing inspections and only includes closed cases. The data includes building addresses, property wards, each fire code violated and a description, and the open and close date of each inspection. In this data, the property address, the property's ward, the fire code violated, and the number of violations at each property were key to uncovering patterns in fire investigations (see @tbl-cleaned_fire_inspection_data). 

```{r}
#| message: false
#| echo: false
#| label: tbl-cleaned_fire_inspection_data
#| tbl-cap: Sample of Cleaned Fire Inspection Data

# Code referenced from: https://github.com/InessaDeAngelis/Toronto_Elections/blob/main/outputs/paper/toronto_elections_paper.qmd

head (cleaned_fire_inspection_data, 4) |>
  kable(
    col.names = c("Property Address", "Property Ward", "Violation Fire Code", "Number of Violations"),
  booktabs = TRUE
  )

```

### Ward Profiles Data
For the second dataset, I used the Ward Profile dataset [@WardProfiles]. This data is gathered from the 2021 Statistics Canada Census of Population and includes demographic, economic, and social information based on the 25 Wards which geographically divide Toronto. While the entire dataset includes data from the 2016 and 2011 Census data, I only used the 2021 data. I focused  on socioeconomic data like median household and individual income, as well as relevant population data like elderly and children populations (see @tbl-ward_rent_data_sample and @tble-ward_socials_data_sample).

```{r}
#| message: false
#| echo: false
#| label: tbl-ward_rent_data_sample
#| tbl-cap: Sample of Cleaned Average Rent Ward Data

# Code referenced from: https://github.com/InessaDeAngelis/Toronto_Elections/blob/main/outputs/paper/toronto_elections_paper.qmd
head (cleaned_ward_data_sample, 6) |>
  kable(
    col.names = c("Data Type", "Ward 1", "Ward 2", "Ward 3", "Ward 4", "Ward 5"),
  booktabs = TRUE
  )
```


```{r}
#| message: false
#| echo: false
#| label: tbl-ward_socials_data_sample
#| tbl-cap: Sample of Elderly Residents

# Code referenced from: https://github.com/InessaDeAngelis/Toronto_Elections/blob/main/outputs/paper/toronto_elections_paper.qmd
head (cleaned_socials_ward_data_sample, 5) |>
  kable(
    col.names = c("Age range", "Ward 1", "Ward 2", "Ward 3", "Ward 4", "Ward 5"),
  booktabs = TRUE
  )
```

### Violations per Ward
The number of violations was highest in Ward 13, at 2662 violations (see @fig-violations_per_ward). Ward 25 had the lowest number of highrise violations at 134. These numbers are most likely influenced by the total number of highrises in the wards since Ward 13 is Toronto Centre and Ward 25 is the Scarborough-Rouge Park area. 

```{r}
#| message: false
#| echo: false
#| label: fig-violations_per_ward
#| fig-cap: Violations per Ward

# Violations per Ward
#Code referenced from: https://github.com/InessaDeAngelis/Toronto_Elections/blob/main/outputs/paper/toronto_elections_paper.qmd

cleaned_fire_inspection_data_ward |>
 ggplot(mapping = aes(x = property_ward, y = number_of_violations_per_ward)) +
  scale_x_continuous(breaks = scales::pretty_breaks(n = 25)) +
  scale_y_continuous(breaks = scales::pretty_breaks(n = 10))+
  geom_bar(stat = "identity") +
  labs(
    title = "Number of Violations per Ward",
    x = "Ward",
    y = "Violations",
  ) +
  theme_classic() 
```

## Appendix
### Data Cleaning
To clean the Fire Investigation data, I removed variables irrelevant to the intended scope of investigation. For example, date of inspections and buildings without fire code violations were removed [@FireInspectionResultsData]. Originally, I wanted to explore the relationship between the number of fire code violations in a ward with socioeconomic data of the same ward, but I had a really hard time adjusting the number of violations  with the number of highrises in the ward, so I had to change and limit the scope of my paper. Attempts to do this can be found at [@attempt]

## References
